---
title: "SplitAndApply"
author: "Aaron Zhang"
date: "2/9/2019"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Goals
- Learn that functions are also objects and can be passed around as an argument
- Learn to use *apply* to apply functions on lists of data

Some of the following materials are adapted from [swcarpentry](https://github.com/swcarpentry/r-novice-gapminder) and [hbctraining](https://github.com/hbctraining/Intro-to-R)

## Summary statistics

In this section, we will introduce the *split* function that is very useful when combined with *apply* for summary statistics.

Question: 
What is the average weight of each species in the surveys data?
We can ask many similar questions that appear extremely frequently:
What is the total weight of all rodents captured at each plot?
What is the max weight of rodents with species_id "AB" captured during each year?

The solutions to the above questions are almost identical: 
* choose a subset of the original dataset for the question and then 
* call an appropriate function.

Intuitively, you might want to write a for loop to select the weights for each species, and then call mean(). This will work and you should try it out.
```{r}
surveys <- read.csv("../surveys.csv")

weight_DM <- surveys[surveys$species_id == "DM",]$weight
mean(weight_DM, na.rm = TRUE)

species_names <- unique(surveys$species_id)
m <- vector(mode = "numeric", length = length(species_names))
for (i in 1:length(species_names)) {
  weights <- surveys[surveys$species_id == species_names[i],]$weight
  m[i] <- mean(weights, na.rm = TRUE)
}
names(m) <- species_names
m
```

There is a more efficient way: using **split** and **apply**. 

## Split
How to split *weight* according to *species_id*?
```{r}
w <- split(surveys$weight, surveys$species_id)
```

## Apply: passing functions as arguments

```
- apply     Apply Functions Over Array Margins
- lapply    Apply a Function over a List or Vector (returns list)
- sapply    Apply a Function over a List or Vector (returns vector)
- mapply    Apply a Function to Multiple List or Vector Arguments
- tapply    Apply a Function Over a Ragged Array
- vapply    Apply a Function over a List or Vector (returns vector with pre-specified type of return value)
- rapply    Recursively Apply a Function to a List
```

The first argument is a list of objects, the second argument is the name of a function, followed by arguments for the second argument. 

```{r}
sapply(w, mean, na.rm = TRUE)
```

There are many R packages that simplify various **apply** calls, a particularly good one being "dplyr" by Hadley Wickham (in the "tidyverse" collection). While learning the functions in base R is still helpful for one to learn the concepts, those packages are more commonly used in practice.

## Exercise

1. Run the following codes, think about what they are trying to achieve.
```{r}
# a)
w <- split(surveys$weight, as.factor(surveys$year))
sapply(w, mean, na.rm = TRUE)

# b)
w <- split(surveys$weight, as.factor(surveys$plot_id))
sapply(w, sum, na.rm = TRUE)

# c)
sapply(split(surveys, surveys$species_id), nrow)
```

Solutions
a) average weights of all animals captured during each year
b) total weight of all animals captured at each plot
c) number of captured animals for each species
