---
title: "Function"
author: "Aaron Zhang"
date: "1/27/2019"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goals
- Describe and utilize functions in R.
- Modify default behavior of functions using arguments in R.
- Describe how R matches arguments
- Practice commonly used functions

Some of the following materials are adapted from [swcarpentry](https://github.com/swcarpentry/r-novice-gapminder) and [hbctraining](https://github.com/hbctraining/Intro-to-R)

# Functions
A key feature of R is functions. Functions are self contained modules of code that accomplish a specific task. Functions usually take in some sort of data structure (value, vector, dataframe etc.), process it, and return a result.

The general usage for a function is the name of the function followed by parentheses:

```r
function_name(input)
```

## Arguments
The input(s) are called arguments, which can include:

* the physical object (any data structure) on which the function carries out a task
* specifications that alter the way the function operates (e.g. options)

In the following example, *sqrt* requires one argument (a) and return the square root.
```{r}
a <- 16
b <- sqrt(a)
b
```

Not all functions take arguments, for example:

```r
getwd()
```

However, most functions can take several arguments. If you don't specify all arguments when calling the function, you will either receive an error or the function will fall back on using a default.

The defaults represent standard values that the author of the function specified as being "good enough in standard cases". However, if you want something specific, simply change the argument yourself with a value of your choice.

An example is the *round()* function: the default value for *digits* is 0. 
```r
round(x, digits = 0)
```

Here, we’ve called round() with just one argument, 3.14159, and it has returned the value 3. That’s because the default is to round to the nearest whole number.
```{r}
round(3.1415926)
```

If we want more digits we can see how to do that by getting information about the round function. We see that if we want a different number of digits, we can type *digits=2* or other decimal places.
```{r}
?round
round(3.1415926, digits = 2)
```

## Argument matching
When a function takes in multiple arguments, R matches the values to the arguments by their names, or by their position. The following calls are identical:
```{r}
# match by name
round(x = 3.1415926, digits = 2)
round(digits = 2, x = 3.1415926)
# match by position
round(3.1415926, 2)
# mix name and position
round(3.1415926, digits = 2)
round(digits = 2, 3.1415926)
# okay but avoid partial name matching
round(3.1415926, d = 2)
round(3.1415926, digi = 2)
```

Being explicit can help improve the readability of your code. 

## Basic functions
We have already used a few examples of basic functions in the previous lessons i.e getwd(), sqrt(), and factor(). These functions are available as part of R's built in capabilities, and we will explore a few more of these base functions below.

```{r}
surveys <- read.csv("../surveys.csv")
lengths <- surveys$hindfoot_length
head(lengths)
tail(lengths)
summary(lengths)
```

We will illustrate some commonly used statistics functions to *lengths*. 
```{r}
sum_lengths <- sum(lengths)
sum_lengths
```

*Alert: Missing values!*
If you have missing values (**NA**), the functions will always return **NA**. The solution is to remove NA by overwriting the default option na.rm from **FALSE** to **TRUE**. 

```{r}
sum_lengths <- sum(lengths, na.rm = TRUE)
sum_lengths
mean_lengths <- mean(lengths, na.rm = TRUE)
mean_lengths
sd_lengths <- sd(lengths, na.rm = TRUE)
sd_lengths
median_lengths <- median(lengths, na.rm = TRUE)
median_lengths
max_lengths <- max(lengths, na.rm = TRUE)
max_lengths
min_lengths <- min(lengths, na.rm = TRUE)
min_lengths
```

## Nested function call
Thus far, to perform any specific task, we have executed every function separately; if we wanted to use the results of a function for downstream purposes, we saved the results to a variable. As you become more comfortable with R, you will find that it is more efficient to code using nested functions, or functions within other functions, which will allow you to execute multiple commands at the same time.

Here is an example. We want to calculate the mean of length, and then round the number to 2 decimal places. The following solutions are equivalent.
```{r}
# solution 1
mean_lengths <- mean(lengths, na.rm = TRUE)
round(mean_lengths, 2)

# solution 2
round(mean(lengths, na.rm = TRUE), 2)
```

Here is another example. Count the number of NA in lengths. *is.na()* will test each element of a vector and return a logical vector of the same length, in which TRUE means the value at that position is NA and FALSE otherwise. We then call *sum()* on the vector to get the count of TRUEs (remember TRUE is 1 and FALSE is 0 in calculation). 
```{r}
sum(is.na(lengths))
```

To improve code readibility, do not chain TOO many function calls!

## Vectorization
Most of R's functions are vectorized, meaning that the function will operate on all elements of a vector without needing to loop through and act on each element one at a time. This makes writing code more concise, easy to read, and less error prone.

For example, if you pass a vector of numbers to sqrt(), it will take square root of each element and return them as a vector.
```{r}
random_numbers <- c(23.7, 16, 5.6, 90, 2)
sqrt(random_numbers)
```

R operators are also functions, and they work on vectors too. When you add a number to a vector, R adds the number to each element of the vector. 
```{r}
x <- 1:4
x * 10
```
The multiplication happened to each element of the vector.

We can also add two vectors together:

```{r}
y <- 5:8
x + y
```
Each element of x was added to its corresponding element of y:
```r
x:   1  2  3  4
     +  +  +  +
y:   5  6  7  8
----------------
     6  8  10 12
```

Comparison operators and logical operators are also vectorized:
```{r}
x > 2
(x > 2) & (x < 4)
```

## Summary
- Functions are modules of codes for certain task. 
- Functions can have 0, 1 or multiple arguments. Some arguments are required, and others are optional with a default value. 
- R matches function arguments by name or position.
- Pay attention to NAs when calling functions. 
- Many functions work on vectors. 

## Exercise

1. The runif() function generates random number within a certain range. Use ?runif to lookup the function, and a) return 10 random numbers with the default options; b) return 9 random numbers between 0.5 and 0.6. 

2. Read in surveys data, extract the weight column and assign it to a variable, then find the sum, mean, sd, min, max of it.

3. We want to know the value t = 1 / (2 \* 3) + 1 / (3 \* 3) + 1 / (4 \* 3) + ... + 1 / (50 * 3). Use vectorization to calculate it (round to the fourth digit)

  



Solutions

Exercise 1
```{r}
# a)
runif(10)
# b)
runif(10, min = 0.5, max = 0.6)
```
Exercise 2
```{r}
surveys <- read.csv("../surveys.csv")
weights <- surveys$weight
sum(weights, na.rm = TRUE)
mean(weights, na.rm = TRUE)
sd(weights, na.rm = TRUE)
median(weights, na.rm = TRUE)
max(weights, na.rm = TRUE)
min(weights, na.rm = TRUE)
```
Exercise 3
```{r}
n <- 2:50
t <- sum(1 / (n * 3))
round(t, 4)
# or
round(sum(1 / ((2:50) * 3)), 4)
```

